Supplementary Appendix

### 2025-07-23 paper/preprint/appendix.tex

\usepackage[title]{appendix}
\usepackage{placeins}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\usepackage{siunitx}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Supplementary Appendix},
  pdfkeywords={Counterfactual Training, Counterfactual
Explanations, Algorithmic Recourse, Explainable AI, Representation
Learning},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training }
\title{Supplementary Appendix}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Counterfactual Training: Teaching Models Plausible and
Actionable Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{}
\date{}
\begin{document}
\maketitle
\begin{abstract}
This is the supplementary appendix to our paper titled
\emph{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}. It provides helpful details on mathematical notations and
formulas, our proposed training regime, extended empirical findings,
hyperparameter tuning and grid searches as well as software and
computations.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Training \sep Counterfactual
Explanations \sep Algorithmic Recourse \sep Explainable AI \sep 
Representation Learning

@@

@@

\begin{itemize}
\tightlist
\item
  \(y^+\): The target class and also the index of the target class.
\item
  \(y^-\): The non-target class and also the index of non-the target
  class.
\item
  \(\mathbf{x}\): a single training sample.
\item
  \(\mathbf{x}^\prime\): a counterfactual.
\item
  \(\mathbf{x}^+\): a training sample in the target class
  (ground-truth).
\item
  \(\mathbf{y}^+\): The one-hot encoded output vector for the target
  class.
\item
  \(\theta\): Model parameters (unspecified).
\item
  \(\Theta\): Matrix of parameters.
\item
  \(\mathbf{M}(\cdot)\): linear predictions (logits) of the classifier.
\end{itemize}

\subsection{Other Technical Details}\label{other-technical-details}

@@

@@

In this section, we provide some background on gradient-based
counterfactual generators (Section~\ref{sec-app-ce-background}) and
discuss how we define convergence in this context
(Section~\ref{sec-app-conv}).

@@

Gradient-based counterfactual search was originally proposed by Wachter,
Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}). It generally solves
the following unconstrained objective,

\[
\begin{aligned}
\min_{\mathbf{z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(\mathbf{M}_{\theta}(g(\mathbf{z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(g(\mathbf{z}^\prime)) }  \right\} 
\end{aligned} 
\]

where \(g: \mathcal{Z} \mapsto \mathcal{X}\) is an invertible function
that maps from the \(L\)-dimensional counterfactual state space to the
feature space and \(\text{cost}(\cdot)\) denotes one or more penalties
that are used to induce certain properties of the counterfactual
outcome. As above, \(\mathbf{y}^+\) denotes the target output and
\(\mathbf{M}_{\theta}(\mathbf{x})\) returns the logit predictions of the
underlying classifier for \(\mathbf{x}=g(\mathbf{z})\).

For all generators used in this work we use standard logit crossentropy
loss for \(\text{yloss}(\cdot)\). All generators also penalize the
distance (\(\ell_1\)-norm) of counterfactuals from their original
factual state. For \emph{Generic} and \emph{ECCCo}, we have
\(\mathcal{Z}:=\mathcal{X}\) and
\(g(\mathbf{z})=g(\mathbf{z})^{-1}=\mathbf{z}\), that is counterfactual
are searched directly in the feature space. Conversely, \emph{REVISE}
traverses the latent space of a variational autoencoder (VAE) fitted to
the training data, where \(g(\cdot)\) corresponds to the decoder
(\citeproc{ref-joshi2019realistic}{Joshi et al. 2019}). In addition to
the distance penalty, \emph{ECCCo} uses an additional penalty component
that regularizes the energy associated with the counterfactual,
\(\mathbf{x}^\prime\) (\citeproc{ref-altmeyer2024faithful}{Altmeyer et
al. 2024}).

@@

In the main paper, we explain that to avoid penalizing implausibility
that arises due to mutability constraints, we impose a point mass prior
on \(p(\mathbf{x})\) for the corresponding feature. We argue that this
approach induces models to be relatively less sensitive to immutable
features, propose a theoretical result supporting this and provide
empirical evidence that strengthens our argument (both in the main paper
and additional findings in this appendix). Below we derive the
analytical results in Prp.\textasciitilde{}\ref{prp-mtblty}.

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-default}: Training Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]

@@

For all of our evaluations, we proceed as follows: for each experiment
setting we generate multiple counterfactuals (``No.~Counterfactuals''),
randomly choosing the factual and target class each time
(Note~\ref{nte-eval-default}). We do this across multiple rounds
(``No.~Runs'') with different random seeds to account for stochasticity
(Note~\ref{nte-eval-default}). This is in line with standard practice in
the related literature on CE. Note~\ref{nte-eval-default} presents the
default hyperparameters used during evaluation. For our final results
presented in the main paper, we rely on held out test sets to sample
factuals (and outputs for our performance metrics). For tuning purposes
we rely on training or validation sets.

\subsubsection{Robust Accuracy}\label{robust-accuracy}

To evaluate robust accuracy (Acc.\(^*\)), we use the Fast Gradient Sign
Method (FGSM) to perturb test samples
(\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens, and Szegedy
2015}). For the main results, we have set the perturbation size to
\(\epsilon=0.03\). We have also tested other perturbation sizes, as well
as randomly perturbed data. Although not reported here, we have
consistently found strong outperformance of CT compared to the weak
baseline.

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-eval-default}: Evaluation Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]

@@

\begin{itemize}
\tightlist
\item
  Counterfactual Parameters:

  \begin{itemize}
  \tightlist
  \item
    Convergence: \texttt{threshold}
  \item
    Decision Threshold: \texttt{0.95}
  \item
    Generator Parameters:

    \begin{itemize}
    \tightlist
    \item
      Decision Threshold: \texttt{0.75}
    \item
      \(\lambda_{\text{cst}}\): \texttt{0.001}
    \item
      \(\lambda_{\text{egy}}\): \texttt{5.0}
    \item
      Learning Rate: \texttt{0.25}
    \item
      Maximum Iterations: \texttt{30}
    \item
      Optimizer: \texttt{sgd}
    \item
      Type: \texttt{ECCo}
    \end{itemize}
  \item
    Maximum Iterations: \texttt{50}
  \item
    No.~Individuals: \texttt{100}
  \item
    No.~Runs: \texttt{5}
  \end{itemize}
\end{itemize}

@@

@@

As discussed the main paper, CT is sensitive to certain hyperparameter
choices. We study the effect of many hyperparameters extensively in
Section~\ref{sec-app-grid}. For the main results, we tune a small set of
key hyperparameters (Section~\ref{sec-app-tune}). The final choices for
the main results are presented for each data set in
Table~\ref{tbl-final-params} along with training, test and batch sizes.

@@

\begin{tabular}{cccccccc}
  \toprule
  \textbf{Data} & \textbf{No. Train} & \textbf{No. Test} & \textbf{Batchsize} & \textbf{Domain} & \textbf{Decision Threshold} & \textbf{No. Counterfactuals} & \textbf{$\lambda_{\text{reg}}$} \\\midrule
  LS & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.01 \\
  Circ & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.5 \\
  Moon & 3600 & 600 & 30 & none & 0.9 & 1000 & 0.25 \\
  OL & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.25 \\\midrule
  Adult & 26049 & 5010 & 1000 & none & 0.75 & 5000 & 0.25 \\
  CH & 16504 & 3101 & 1000 & none & 0.5 & 5000 & 0.25 \\
  Cred & 10617 & 1923 & 1000 & none & 0.5 & 5000 & 0.25 \\
  GMSC & 13371 & 2474 & 1000 & none & 0.5 & 5000 & 0.5 \\
  MNIST & 11000 & 2000 & 1000 & (-1.0, 1.0) & 0.5 & 5000 & 0.01 \\
\end{tabular}

@@

\subsection{Final Results}\label{sec-app-final-results}

Plus/minus two standard deviations of bootstrap estimates.

\subsubsection{Robust Performance Plots}\label{robust-performance-plots}

@@

@@

\caption{\label{tbl-ci}Mean outcomes for \textbf{CT} and \textbf{BL}
along with bootstrapped confidence intervals (99\%) for difference in
mean outcomes grouped by dataset and evaluation metric. Column
\textbf{LB} and \textbf{UB} show the lower and upper bound of the
intervals, respectively, and computed using the percentile method. The
underlying counterfactual evaluations are the same as the ones used to
produce the main table in the paper.}

@@

\begin{tabular}{
l
S[table-format=2.2(3.2)]
S[table-format=2.2(2.2)]
}
  \toprule
  \textbf{Data} & \textbf{CT} & \textbf{BL} \\\midrule
  LS & 0.03 \pm 0.0 & 10.24 \pm 2.4 \\
  Circ & 3.2 \pm 0.67 & 149.76 \pm 842.75 \\
  Moon & 60.84 \pm 128.51 & 0.55 \pm 0.06 \\
  OL & 0.78 \pm 0.12 & 4.81 \pm 1.08 \\\midrule
  Adult & 0.43 \pm 0.01 & 1.0 \pm 0.0 \\
  CH & 0.08 \pm 0.01 & 0.23 \pm 0.01 \\
  Cred & 0.0 \pm 0.0 & 0.43 \pm 0.01 \\
  GMSC & 1.0 \pm 0.0 & 0.21 \pm 0.03 \\
  MNIST & 0.18 \pm 0.01 & 0.41 \pm 0.01 \\
\end{tabular}

@@

@@

\begin{minipage}{0.33\linewidth}

\subcaption{\label{tbl-costs}Costs}

@@

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\subcaption{\label{tbl-val}Validity}

@@

\begin{tabular}{lcc}
  \toprule
  \textbf{Data} & \textbf{CT} & \textbf{BL} \\\midrule
  LS & 1.0 & 1.0 \\
  Circ & 0.97 & 0.52 \\
  Moon & 1.0 & 1.0 \\
  OL & 0.87 & 0.98 \\\midrule
  Adult & 0.61 & 0.99 \\
  CH & 0.96 & 1.0 \\
  Cred & 0.7 & 1.0 \\
  GMSC & 0.63 & 1.0 \\
  MNIST & 1.0 & 1.0 \\
\end{tabular}

@@

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\subcaption{\label{tbl-val-mtbl}Validity}

>>>>

\usepackage[title]{appendix}
\usepackage{placeins}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{siunitx}
\sisetup{uncertainty-mode = separate}
\DeclareMathSizes{10}{9}{7}{6.5}
\usepackage{enumitem}
\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\usepackage{siunitx}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Supplementary Appendix},
  pdfkeywords={Counterfactual Training, Counterfactual
Explanations, Algorithmic Recourse, Explainable AI, Representation
Learning},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\newcommand{\runninghead}{}
\renewcommand{\runninghead}{Counterfactual Training }
\title{Supplementary Appendix}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Counterfactual Training: Teaching Models Plausible and
Actionable Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{}
\date{}
\begin{document}
\maketitle
\begin{abstract}
This is the supplementary appendix to our paper titled
\emph{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}. It provides helpful details on mathematical notations and
formulas, our proposed training regime, extended empirical findings,
hyperparameter tuning and grid searches as well as software and
computations.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Training \sep Counterfactual
Explanations \sep Algorithmic Recourse \sep Explainable AI \sep 
Representation Learning

@@

\subsection{Variables and Parameters}\label{variables-and-parameters}

@@

\begin{itemize}
\tightlist
\item
  \(\mathcal{Y}\): The output domain.
\item
  \(y^+\): The target class and also the index of the target class.
\item
  \(y^-\): The non-target class and also the index of non-the target
  class.
\item
  \(\mathcal{X}\): The input domain.
\item
  \(\mathbf{x}\): a single training sample.
\item
  \(\mathbf{x}^\prime\): a counterfactual.
\item
  \(t=1,...,T\): Step indicator for counterfactual search iterations.
\item
  \(\mathbf{x}^\prime_{t \leq T}\): a nascent counterfactual, defined as
  a counterfactual that has not yet converged.
\item
  \(\mathbf{x}^\prime_{T}\): a mature counterfactual, defined as a
  counterfactual that has either passed a threshold probability or
  exhausted all \(T\) steps.
\item
  \(\mathbf{x}^+\): a training sample in the target class
  (ground-truth).
\item
  \(\mathbf{y}^+\): The one-hot encoded output vector for the target
  class.
\item
  \(\theta\): Model parameters (unspecified).
\item
  \(\Theta\): Matrix of parameters.
\item
  \(\mathbf{M}(\cdot)\): linear predictions (logits) of the classifier.
\end{itemize}

\subsection{Formulas}\label{formulas}

\subsubsection{Maximum Mean Discrepancy}\label{maximum-mean-discrepancy}

@@

\subsection{Other Conventions}\label{other-conventions}

In some place of this appendix, we use the terms \emph{full/Full}
(i.e.~the full CT objective) and \emph{vanilla/Vanilla} (i.e.~vanilla
training objective) to refer to models trained with counterfactual
training (\emph{CT}) and the baseline (\emph{BL}), respectively.

@@

In this section, we provide some additional background on gradient-based
counterfactual generators (Section~\ref{sec-app-ce-background}) and
discuss how we define convergence in this context
(Section~\ref{sec-app-conv}).

@@

Gradient-based counterfactual search was originally proposed by Wachter,
Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}). A more general version
of the counterfactual search objective presented in the paper is as
follows,

\[
\begin{aligned}
\min_{\mathbf{z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(\mathbf{M}_{\theta}(g(\mathbf{z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{reg}(g(\mathbf{z}^\prime)) }  \right\} 
\end{aligned} 
\]

where \(g: \mathcal{Z} \mapsto \mathcal{X}\) is an invertible function
that maps from the \(L\)-dimensional counterfactual state space to the
feature space and \(\text{reg}(\cdot)\) denotes one or more penalties
that are used to induce certain properties of the counterfactual
outcome. As above, \(\mathbf{y}^+\) denotes the target output and
\(\mathbf{M}_{\theta}(\mathbf{x})\) returns the logit predictions of the
underlying classifier for \(\mathbf{x}=g(\mathbf{z})\).

For all generators used in this work we use standard logit crossentropy
loss for \(\text{yloss}(\cdot)\). All generators also penalize the
distance (\(\ell_1\)-norm) of counterfactuals from their original
factual state. For \emph{Generic} and \emph{ECCCo}, we have
\(\mathcal{Z}:=\mathcal{X}\) and
\(g(\mathbf{z})=g(\mathbf{z})^{-1}=\mathbf{z}\), that is counterfactual
are searched directly in the feature space. Conversely, \emph{REVISE}
traverses the latent space of a variational autoencoder (VAE) fitted to
the training data, where \(g(\cdot)\) corresponds to the decoder
(\citeproc{ref-joshi2019realistic}{Joshi et al. 2019}). In addition to
the distance penalty, \emph{ECCCo} uses a penalty that regularizes the
energy associated with the counterfactual, \(\mathbf{x}^\prime\)
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). We omit the
conformal set size penalty proposed in the original paper, since a) the
authors found faithfulness to primarily depend on the energy penalty and
hence this alleviates us from one additional hyperparameter.

@@

In the main paper, we explain that to avoid penalizing implausibility
that arises due to mutability constraints, we impose a point mass prior
on \(p(\mathbf{x})\) for the corresponding feature. We argue that this
approach induces models to be relatively less sensitive to immutable
features, propose a theoretical result supporting this and provide
empirical evidence that strengthens our argument (both in the main paper
and additional findings in this appendix). Below we derive the
analytical results in Proposition in the main paper.

@@

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-train-default}: Training Phase}]

@@

\subsubsection{Counterfactual Outcomes}\label{counterfactual-outcomes}

For all of our counterfactual evaluations, we proceed as follows: for
each dataset we run \(J\) bootstrap rounds (``No.~Runs'') to account for
stochasticity (Note~\ref{nte-eval-default}); for each bootstrap round,
we randomly draw factual and target pairs; then, for each model, we draw
samples from the test set (with replacement) for which the model
predicts the randomly chosen factual class; finally, we generate
multiple counterfactuals (``No.~Counterfactuals'') and evaluate the
outcomes (Note~\ref{nte-eval-default}). This is in line with standard
practice in the related literature on CE. For our final results
presented in the main paper, we rely on held-out test sets for
evaluation. For tuning purposes we rely on training and/or validation
sets.

Note~\ref{nte-eval-default} presents the default hyperparameters used
during evaluation for tuning purposes. For the main results presented in
the paper, we use larger evaluations, specifically:

\begin{itemize}
\tightlist
\item
  ``No.~Runs'': We set the number of bootstrap rounds to \(J=100\) for
  all datasets.
\item
  ``No.~Individuals'': In each round we draw 1,250, 500 and 125 samples
  for synthetic datasets, real-world tabular datasets and \emph{MNIST},
  respectively, across five different values for the strength of the
  energy penalty of \emph{ECCCo} at test time,
  \(\lambda_{\text{egy}}\in\{0.1, 0.5, 1.0, 5.0, 10.0\}\).
\end{itemize}

\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}]

@@

\begin{itemize}
\tightlist
\item
  Convergence: \texttt{threshold}
\item
  Decision Threshold: \texttt{0.95}
\item
  Maximum Iterations: \texttt{50}
\item
  No.~Individuals: \texttt{100}
\item
  No.~Runs: \texttt{5}
\end{itemize}

@@

To assess (robust) predictive performance, we evaluate model accuracy on
(adversarially perturbed) test data. To generate adversarial examples we
use the Fast Gradient Sign Method (FGSM)
(\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens, and Szegedy
2015}). For the main results in the paper, we choose a range of values
\(\epsilon=[0.0,0.1]\). In some places of this appendix, you will also
find predictive performance evaluations in terms of the F1-score.

@@

As discussed the main paper, CT is sensitive to certain hyperparameter
choices. We study the effect of many hyperparameters extensively in
Section~\ref{sec-app-grid} of this appendix. For the main results, we
tune a small set of key hyperparameters (Section~\ref{sec-app-tune}).
The final choices for the main results are presented for each data set
in Table~\ref{tbl-final-params} along with training, test and batch
sizes.

@@

\begin{tabular}{cccccccc}
  \toprule
  \textbf{Data} & \textbf{No. Train} & \textbf{No. Test} & \textbf{Batchsize} & \textbf{Domain} & \textbf{Decision Threshold} & \textbf{No. Counterfactuals} & \textbf{$\lambda_{\text{reg}}$} \\\midrule
  LS & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.01 \\
  Circ & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.5 \\
  Moon & 3600 & 600 & 30 & none & 0.9 & 1000 & 0.25 \\
  OL & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.25 \\\midrule
  Adult & 26049 & 5010 & 1000 & none & 0.75 & 5000 & 0.25 \\
  CH & 16504 & 3101 & 1000 & none & 0.5 & 5000 & 0.25 \\
  Cred & 10617 & 1923 & 1000 & none & 0.5 & 5000 & 0.25 \\
  GMSC & 13371 & 2474 & 1000 & none & 0.5 & 5000 & 0.5 \\
  MNIST & 11000 & 2000 & 1000 & (-1.0, 1.0) & 0.5 & 5000 & 0.01 \\\bottomrule
\end{tabular}

@@

@@

Table~\ref{tbl-ci} present the exact confidence intervals (99\%) for the
difference in mean outcomes on which we base our assessment of
statistical significance in the main paper. Grouped by evaluation
metrics (Variable) and dataset (Data), the table presents the mean
outcomes for CT and BT and finally the lower bound (LB) and upper bound
(UB) of the confidence interval. To compute the intervals, we used the
percentile method for bootstrapped confidence intervals: the lower and
upper bound represent the \(\alpha/2\)- and \((1-\alpha / 2)-\)quantile
of the bootstrap distribution, respectively, for \(\alpha=0.01\).

@@

\caption{\label{tbl-ci}Mean outcomes for CT and BL along with
bootstrapped confidence intervals (99\%) for difference in mean outcomes
grouped by dataset and evaluation metric. Column LB and UB show the
lower and upper bound of the intervals, respectively, and computed using
the percentile method. The underlying counterfactual evaluations are the
same as the ones used to produce the main table in the paper.}

@@

\begin{tabular}{
l
S[table-format=2.2(3.2)]
S[table-format=2.2(2.2)]
}
  \toprule
  \textbf{Data} & \textbf{CT} & \textbf{BL} \\\midrule
  LS & 0.03 \pm 0.0 & 10.24 \pm 2.4 \\
  Circ & 3.2 \pm 0.67 & 149.76 \pm 842.75 \\
  Moon & 60.84 \pm 128.51 & 0.55 \pm 0.06 \\
  OL & 0.78 \pm 0.12 & 4.81 \pm 1.08 \\\midrule
  Adult & 0.43 \pm 0.01 & 1.0 \pm 0.0 \\
  CH & 0.08 \pm 0.01 & 0.23 \pm 0.01 \\
  Cred & 0.0 \pm 0.0 & 0.43 \pm 0.01 \\
  GMSC & 1.0 \pm 0.0 & 0.21 \pm 0.03 \\
  MNIST & 0.18 \pm 0.01 & 0.41 \pm 0.01 \\\bottomrule
\end{tabular}

@@

In Table~\ref{tbl-panel}, we present additional outcomes for common
evaluation metrics: Table~\ref{tbl-costs} presents the average reduction
in costs of counterfactuals for CT vs.~BL with no mutability
constraints, i.e.~corresponding to the first two columns in the main
table of the paper; Table~\ref{tbl-val} shows the corresponding average
validities; finally, Table~\ref{tbl-val-mtbl} shows average validities
for the case with mutability constraints, i.e.~corresponding to the
third columns in the main table of the paper.

As noted in the discussion section of the main paper, we observe mixed
results results here. Average costs in terms of distances from factual
values decrease for most datasets, which is positively surprising since
improved plausibility requires counterfactuals to travel further into
the target domain than minimum distance counterfactuals. It appears that
in these cases faithful counterfactuals for the baseline model still end
up far away from their initial starting points, but not close enough for
samples in the target domain to be plausible. In that sense, CT can be
seen to improve both plausibility and costs for faithful CE. In some
cases though (\emph{LS}, \emph{CH}, \emph{MNIST}), we do seem to observe
the tradeoff between plausibility and costs play out, as we would expect
(compare panels (a) and (b) of Figure 1 in the main paper for
reference).

Concerning validity, we find that can lead to substantial reductions and
only increases average validity compared to the baseline in one case
(\emph{Circ}). As noted in the discussion section of the main paper,
this result does not surprise us: by design, CT shrinks the solution
space for valid counterfactual explanations, thus making it ``harder''
to reach validity compared to the baseline model. Note that for a number
of reasons this should not be seen as problematic:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Validity of gradient-based CE is a function on the number of steps and
  the step size which we both kept fixed during evaluation: simply
  adjusting \(T=50\) to higher values or choosing a larger step size
  will lead to higher rates of validity.
\item
  Even though reaching validity is sometimes ``harder'' in terms of the
  necessary number of steps for a given step size, we have already shown
  that the average distances that counterfactuals need to travel
  decrease for most datasets. Users care about costs in terms of feature
  distances, not search iteration steps.
\item
  From a philosophical perspective on algorithmic recourse, validity in
  and off itself is not a sufficient desideratum for counterfactuals. In
  fact, Venkatasubramanian and Alfano
  (\citeproc{ref-ven2020philosophical}{2020}) propose introducing an
  upper bound on costs of the flipset (i.e.~the set of valid CE),
  arguing that valid but highly costly counterfactuals are not useful to
  individuals in practice. In a similar fashion, it could be argued that
  there should be an upper bound on the implausibility of
  counterfactuals in the flipset.
\end{enumerate}

@@

\begin{minipage}{0.40\linewidth}

\subcaption{\label{tbl-costs}Reduction in average costs for CT vs.~the
baseline. Results correspond to the case with no mutability constraints
in the main table of the paper.}

@@

\end{minipage}%
%
\begin{minipage}{0.20\linewidth}
~\end{minipage}%
%
\begin{minipage}{0.40\linewidth}

\subcaption{\label{tbl-val}Average validities of counterfactuals for CT
and BL. Unconstrained case.}

@@

\begin{tabular}{lcc}
  \toprule
  \textbf{Data} & \textbf{CT} & \textbf{BL} \\\midrule
  LS & 1.0 & 1.0 \\
  Circ & 0.97 & 0.52 \\
  Moon & 1.0 & 1.0 \\
  OL & 0.87 & 0.98 \\\midrule
  Adult & 0.61 & 0.99 \\
  CH & 0.96 & 1.0 \\
  Cred & 0.7 & 1.0 \\
  GMSC & 0.63 & 1.0 \\
  MNIST & 1.0 & 1.0 \\\bottomrule
\end{tabular}

@@

\end{minipage}%
\newline
\begin{minipage}{\linewidth}

\subcaption{\label{tbl-val-mtbl}Average validities of counterfactuals
for CT and BL. Mutability constrained case.}
